---
title: "Plasma p-Tau181 and GFAP track 7T MR-based changes in Alzheimer’s disease"
author: "Laura Göschel"
date: "29.11.2023"
output:
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    theme: paper
    code_folding: hide
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{css, echo = FALSE}
caption {
      color: black;
    }
```

```{r library setup, include=FALSE}
library(readxl)
library(gtsummary)
library(tidyverse)
library(ggpubr)
library(flextable)
library(kableExtra)
library(ggeffects)
library(lme4)
library(emmeans)
library(finalfit)
library(patchwork)
library(knitr)
library(ggtext)
library(Ckmeans.1d.dp)
```

# Statistical Analysis Code Disclaimer

 This R script contains the code used for the statistical analysis of the research paper titled
 "Plasma p-Tau181 and GFAP track 7T MR-based changes in Alzheimer’s disease." The purpose of
 sharing this code is to provide transparency and facilitate reproducibility of the reported
 results. Please note the following:

 1. I am not a professional statistician or IT expert. The code provided here may have flaws
    or suboptimal practices.

 2. Use this code at your own risk. I make no guarantees about the correctness or reliability
    of the code.

 3. Feel free to use, modify, and distribute this code for educational and research purposes,
    but be aware that it comes with no warranty or support.

 4. If you encounter issues or have suggestions for improvement, please consider contributing
    by submitting a pull request or opening an issue on the GitHub repository.

 5. This code is provided "as is", and I disclaim any responsibility or liability for any
    errors, inaccuracies, or consequences arising from the use of this code.

Thank you for your understanding and collaboration in promoting open and transparent science.

```{r Data preparation, include = FALSE}
NeuroMET <- read_excel("yourdata.xlsx", 
    sheet = "yourdatasheet")


NeuroMET$diagnose <- factor(NeuroMET$diagnose, labels=c('HC', 'SCD', 'MCI','AD'))
NeuroMET <- NeuroMET %>% filter(!is.na(diagnose))
NeuroMET <- NeuroMET %>% filter(!(Visit == 'smartage'))
NeuroMET$Visit <- factor(NeuroMET$Visit, labels=c('t1', 't2','t3', 't4', 't5'))

NeuroMET$APOE <- factor(NeuroMET$APOE, labels=c('22','23','24','33','34','44'))
NeuroMET$APOEe4 <- ifelse(NeuroMET$APOE=="22" | NeuroMET$APOE=="23" | NeuroMET$APOE=="33", "0", "1")
NeuroMET$APOEe4 <- factor(NeuroMET$APOEe4, labels=c('non-carrier','carrier'))

NeuroMET$edu <- as.numeric(NeuroMET$education_years_cerad)

#global cognition
NeuroMET$mmse <- as.numeric(NeuroMET$mmst_sum)

#memory + executive function 
NeuroMET$nmm <- as.numeric(NeuroMET$NMM_value)
NeuroMET$rwt_food_count <- as.numeric(NeuroMET$rwt_food_count)

#calulate z-scores
NeuroMET$cer_vf_anim_count[NeuroMET$cer_vf_anim_count == "NaN"] <- NA

sum <- NeuroMET %>%
  filter(diagnose == 'HC'& Visit  == 't1')%>%
  dplyr::summarise(mean_animals = mean(cer_vf_anim_count, na.rm = TRUE), sd_animals = sd(cer_vf_anim_count, na.rm = TRUE),
            mean_s = mean(cer_vf_s_count, na.rm = TRUE), sd_s = sd(cer_vf_s_count, na.rm = TRUE),
            mean_digitsymbol = mean(ds_total_score, na.rm = TRUE), sd_digitsymbol = sd(ds_total_score, na.rm = TRUE),
            mean_stroop = mean(stroop_time, na.rm = TRUE), sd_stroop = sd(stroop_time, na.rm = TRUE),
            mean_tmtb = mean(cer_tmt_b1, na.rm = TRUE), sd_tmtb = sd(cer_tmt_b1, na.rm = TRUE),
            mean_mmse = mean(mmse, na.rm = TRUE), sd_mmse = sd(mmse, na.rm = TRUE),
            mean_nmm = mean(nmm, na.rm = TRUE), sd_nmm = sd(nmm, na.rm = TRUE)
  )
NeuroMET$z_score_animals <- (NeuroMET$cer_vf_anim_count - sum$mean_animals) / sum$sd_animals
NeuroMET$z_score_s <- (NeuroMET$cer_vf_s_count - sum$mean_s) / sum$sd_s
NeuroMET$z_score_digitsymbol <- (NeuroMET$ds_total_score - sum$mean_digitsymbol) / sum$sd_digitsymbol
NeuroMET$z_score_stroop <- (NeuroMET$stroop_time - sum$mean_stroop) / sum$sd_stroop
NeuroMET$z_score_stroop_inv <- -1*(NeuroMET$z_score_stroop)
NeuroMET$z_score_tmtb <- (NeuroMET$cer_tmt_b1 - sum$mean_tmtb) / sum$sd_tmtb
NeuroMET$z_score_tmtb_inv <- -1*(NeuroMET$z_score_tmtb)
NeuroMET$z_score_mmse <- (NeuroMET$mmse - sum$mean_mmse) / sum$sd_mmse
NeuroMET$z_score_nmm <- (NeuroMET$nmm - sum$mean_nmm) / sum$sd_nmm

#calculate executive function if not more than 3 items are missing
NeuroMET <- NeuroMET %>% 
  select(record_id, Visit, z_score_animals, z_score_s, z_score_digitsymbol, z_score_stroop_inv, z_score_tmtb_inv) %>%
  filter(rowSums(is.na(.)) <= 2) %>%
  rowwise() %>%
  mutate(exefct = mean(c(z_score_animals, z_score_s, z_score_digitsymbol, z_score_stroop_inv, z_score_tmtb_inv), na.rm = TRUE)) %>%
  ungroup() %>%
  select(record_id, Visit,exefct) %>%
  full_join(NeuroMET, by = c("record_id", "Visit"))

#biomarkers
NeuroMET$Ab40 <- as.numeric(NeuroMET$`Ab40_pg/mL_VUmc`)
NeuroMET$Ab42 <- as.numeric(NeuroMET$`Ab42_pg/mL_Vumc`)
NeuroMET$Ab42_40 <- NeuroMET$Ab42/NeuroMET$Ab40
NeuroMET$pTau <- as.numeric(NeuroMET$`pTau181_pg/mL_Vumc`)
NeuroMET$GFAP <- as.numeric(NeuroMET$`GFAP_pg/mL_Vumc`)
NeuroMET$nfl <- as.numeric(NeuroMET$`NfL_pg/mL_Vumc`)

NeuroMET$Myo <- as.numeric(NeuroMET$`Ins Conc`)
NeuroMET$Myo[NeuroMET$Myo == "NaN"] <- NA
NeuroMET$NAA <- as.numeric(NeuroMET$`NAA Conc`)
NeuroMET$NAA[NeuroMET$NAA == "NaN"] <- NA

# For weighted analyses of MRS data, CRLB are mutated to weights (mean = 1)
NeuroMET$Ins_aCRLB <- as.numeric(NeuroMET$`Ins CRLB abs`) # absolute CRLB
NeuroMET$Ins_inv_aCRLB <- 1/((NeuroMET$Ins_aCRLB)^2) # reciprocal and squared aCRLB
NeuroMET$Ins_CRLB <- ((NeuroMET$Ins_inv_aCRLB - mean(NeuroMET$Ins_inv_aCRLB, na.rm =TRUE))/ (max(NeuroMET$Ins_inv_aCRLB, na.rm =TRUE) - min(NeuroMET$Ins_inv_aCRLB, na.rm =TRUE)))+1
mean(NeuroMET$Ins_CRLB, na.rm = TRUE)   # should be 1 
sum(NeuroMET$Ins_CRLB, na.rm=TRUE)      # should be number of participants

NeuroMET$NAA_aCRLB <- as.numeric(NeuroMET$`NAA CRLB abs`) # absolute CRLB
NeuroMET$NAA_inv_aCRLB <- 1/((NeuroMET$NAA_aCRLB)^2) # reciprocal and squared aCRLB
NeuroMET$NAA_CRLB <- ((NeuroMET$NAA_inv_aCRLB - mean(NeuroMET$NAA_inv_aCRLB, na.rm =TRUE))/ (max(NeuroMET$NAA_inv_aCRLB, na.rm =TRUE) - min(NeuroMET$NAA_inv_aCRLB, na.rm =TRUE)))+1
mean(NeuroMET$NAA_CRLB, na.rm = TRUE)   # should be 1 
sum(NeuroMET$NAA_CRLB, na.rm=TRUE)      # should be number of participants

# Exclude MRI ROI volumes with insufficient quality 
NeuroMET <- NeuroMET %>% mutate(hip_lh_cs= Whole_hippocampus_lh_cs * QC_Hippocampus_subf_lh_cs,
                                hip_rh_cs = Whole_hippocampus_rh_cs * QC_Hippocampus_subf_rh_cs,
                                amyg_lh_cs= Whole_amygdala_lh_cs * QC_Amygdala_subf_lh_cs,
                                amyg_rh_cs=Whole_amygdala_rh_cs * QC_Amygdala_subf_rh_cs,
                                etiv_cs= EstimatedTotalIntraCranialVol_cs,
                                hip_lh_long= left_Whole_hippocampus_long * QC_lh_Hip_long,
                                hip_rh_long= right_Whole_hippocampus_long * QC_rh_Hip_long,
                                amyg_lh_long= left_Whole_amygdala_long * QC_lh_Amygdala_long,
                                amyg_rh_long=right_Whole_amygdala_long * QC_rh_Amygdala_long,
                                parietal_thick_lh_cs = lh_Parietal_lobe_and_PCC_thickness_mean_cs * QC_parietal_ctx_lh_cs,
                                parietal_thick_rh_cs = rh_Parietal_lobe_and_PCC_thickness_mean_cs * QC_parietal_ctx_rh_cs,
                                parietal_thick_lh_long = lh_Parietal_lobe_and_PCC_thickness_long * QC_lh_PCC_long,
                                parietal_thick_rh_long = rh_Parietal_lobe_and_PCC_thickness_long * QC_rh_PCC_long,
                                etiv_long= EstimatedTotalIntraCranialVol_long) %>% 
                         mutate(hip_lh_cs = replace(hip_lh_cs, hip_lh_cs==0, NA),
                                hip_rh_cs = replace(hip_rh_cs, hip_rh_cs==0, NA),
                                amyg_lh_cs = replace(amyg_lh_cs, amyg_lh_cs==0, NA),
                                amyg_rh_cs = replace(amyg_rh_cs, amyg_rh_cs==0, NA),
                                etiv_cs = replace(etiv_cs, etiv_cs==0, NA),
                                hip_lh_long = replace(hip_lh_long, hip_lh_long==0, NA),
                                hip_rh_long = replace(hip_rh_long, hip_rh_long==0, NA),
                                amyg_lh_long = replace(amyg_lh_long, amyg_lh_long==0, NA),
                                amyg_rh_long = replace(amyg_rh_long, amyg_rh_long==0, NA),
                                parietal_thick_lh_cs = replace(parietal_thick_lh_cs, parietal_thick_lh_cs==0, NA),
                                parietal_thick_rh_cs = replace(parietal_thick_rh_cs, parietal_thick_rh_cs==0, NA),
                                parietal_thick_lh_long = replace(parietal_thick_lh_long, parietal_thick_lh_long==0, NA),
                                parietal_thick_rh_long = replace(parietal_thick_rh_long, parietal_thick_rh_long==0, NA),
                                etiv_long = replace(etiv_long, etiv_long==0, NA)
                                )

# Calculate mean or copy unilateral volume data if other side did not pass quality control
mean_lh_rh <- function (lh, rh){
  if (! is.na(lh)){
    if (! is.na(rh)) {
      (lh + rh ) /2
    }
    else if(is.na(rh)){
      (lh + lh ) /2
    }
  }
  else if (is.na(lh)){
    if (! is.na(rh)){
      (rh + rh ) /2
    }
  } else {NA}
}

NeuroMET <- NeuroMET %>%
  mutate(Whole_hip_cs = mapply(mean_lh_rh, hip_lh_cs, hip_rh_cs),
         Whole_amyg_cs = mapply(mean_lh_rh, amyg_lh_cs, amyg_rh_cs),
         Whole_hip_long = mapply(mean_lh_rh, hip_lh_long, hip_rh_long),
         Whole_amyg_long = mapply(mean_lh_rh, amyg_lh_long, amyg_rh_long),
         Whole_parietal_thick_cs = mapply(mean_lh_rh, parietal_thick_lh_cs, parietal_thick_rh_cs),
         Whole_parietal_thick_long = mapply(mean_lh_rh, parietal_thick_lh_long, parietal_thick_rh_long)
         )

# mapply gives out lists with NULLs instead of NA. Replace NULL by NA and unlist volumetric data
NeuroMET$Whole_hip_cs[sapply(NeuroMET$Whole_hip_cs, is.null)] <- NA
NeuroMET$Whole_amyg_cs[sapply(NeuroMET$Whole_amyg_cs, is.null)] <- NA
NeuroMET$Whole_hip_long[sapply(NeuroMET$Whole_hip_long, is.null)] <- NA
NeuroMET$Whole_amyg_long[sapply(NeuroMET$Whole_amyg_long, is.null)] <- NA
NeuroMET$Whole_parietal_thick_cs[sapply(NeuroMET$Whole_parietal_thick_cs, is.null)] <- NA
NeuroMET$Whole_parietal_thick_long[sapply(NeuroMET$Whole_parietal_thick_long, is.null)] <- NA


NeuroMET <- NeuroMET %>%
  mutate(Whole_hip_cs = unlist(Whole_hip_cs),
         Whole_amyg_cs = unlist(Whole_amyg_cs),
         Whole_hip_long = unlist(Whole_hip_long),
         Whole_amyg_long = unlist(Whole_amyg_long),
         Whole_parietal_thick_cs = unlist(Whole_parietal_thick_cs),
         Whole_parietal_thick_long = unlist(Whole_parietal_thick_long)
         )

#remove unwanted variables of volume measures
NeuroMET<- NeuroMET%>% select(-Whole_hippocampus_lh_cs, -Whole_hippocampus_rh_cs,-Whole_amygdala_lh_cs, -Whole_amygdala_rh_cs,
                              -left_Whole_hippocampus_long, -right_Whole_hippocampus_long, -left_Whole_amygdala_long, -right_Whole_amygdala_long,
                              -QC_Hippocampus_subf_lh_cs, -QC_Hippocampus_subf_rh_cs, -QC_Amygdala_subf_lh_cs, -QC_Amygdala_subf_rh_cs,
                              -QC_lh_Hip_long, -QC_rh_Hip_long, -QC_lh_Amygdala_long, -QC_rh_Amygdala_long,
                              -hip_lh_cs,-hip_rh_cs, -amyg_lh_cs,-amyg_rh_cs,-etiv_cs, -hip_lh_long, -hip_rh_long, -amyg_lh_long, -amyg_rh_long,
                              -parietal_thick_lh_cs, -parietal_thick_rh_cs, -parietal_thick_lh_long, -parietal_thick_rh_long, -etiv_long)

#adjust to eTIV cross-sectional (adjusted to HC in T1 only)
ROI_list <- list("Whole_hip_cs", "Whole_amyg_cs")
NeuroMET_HC_t1 <- NeuroMET %>%
  filter (diagnose=="HC" & Visit =="t1")
meanTIV_HC = mean(NeuroMET_HC_t1$EstimatedTotalIntraCranialVol_cs, na.rm = TRUE)
adj_ROIs <- sapply(ROI_list, function (ROI) {
  NeuroMET_HC_t1$ROI_HC <- unlist(NeuroMET_HC_t1[, ROI])
  regr <- lm(ROI_HC ~ EstimatedTotalIntraCranialVol_cs, NeuroMET_HC_t1)
  slope <- summary(regr)$coefficients[2,1]
  adjVol = unlist(NeuroMET[, ROI]) - slope * (NeuroMET$EstimatedTotalIntraCranialVol_cs - meanTIV_HC)
})
adj_ROIs_cs <- data.frame(record_id = NeuroMET$record_id, Visit = NeuroMET$Visit, as.data.frame(adj_ROIs))%>%
  mutate(Whole_hip_cs_adj = V1, Whole_amyg_cs_adj = V2)%>%
  select(-V1, -V2)

#adjust to eTIV longitudinal (adjusted to HC in T1 only)
ROI_list <- list("Whole_hip_long", "Whole_amyg_long")
NeuroMET_HC_t1 <- NeuroMET %>%
  filter (diagnose=="HC" & Visit =="t1")
meanTIV_HC = mean(NeuroMET_HC_t1$EstimatedTotalIntraCranialVol_long, na.rm = TRUE)
adj_ROIs <- sapply(ROI_list, function (ROI) {
  NeuroMET_HC_t1$ROI_HC <- unlist(NeuroMET_HC_t1[, ROI])
  regr <- lm(ROI_HC ~ EstimatedTotalIntraCranialVol_long, NeuroMET_HC_t1)
  slope <- summary(regr)$coefficients[2,1]
  adjVol = unlist(NeuroMET[, ROI]) - slope * (NeuroMET$EstimatedTotalIntraCranialVol_long - meanTIV_HC)
})
adj_ROIs_long <- data.frame(record_id = NeuroMET$record_id, Visit = NeuroMET$Visit, as.data.frame(adj_ROIs)) %>%
  mutate(Whole_hip_long_adj = V1, Whole_amyg_long_adj = V2)%>%
  select(-V1, -V2)

NeuroMET <- NeuroMET%>%
  full_join(adj_ROIs_cs, by = c("record_id", "Visit"))%>%
  full_join(adj_ROIs_long, by = c("record_id", "Visit"))%>%
  select(-Whole_hip_cs, -Whole_amyg_cs, -Whole_hip_long, -Whole_amyg_long) %>% 
  # merge cross sectional and longitudinal data. When there is no Whole_hip_long_adj, then select cross-sectional data 
  mutate(Whole_hip_cs_long = ifelse(is.na(Whole_hip_cs_adj) != is.na(Whole_hip_long_adj), Whole_hip_cs_adj, Whole_hip_long_adj),
         parietal_thick_cs_long = ifelse(is.na(Whole_parietal_thick_cs) != is.na(Whole_parietal_thick_long), Whole_parietal_thick_cs, Whole_parietal_thick_long)) 

#Connectivities. 
NeuroMET <- NeuroMET%>%
  mutate(DMN = Conn_Default,
         Sal = Conn_SalVentAttn)

# exclude this data because of misregistration 
NeuroMET<- NeuroMET %>% mutate(DMN = ifelse(NeuroMET$record_id == "NeuroMet080" & NeuroMET$Visit == "t3" | NeuroMET$record_id == "NeuroMet118" & NeuroMET$Visit == "t2", NA, DMN)) 
NeuroMET<- NeuroMET %>% mutate(Sal = ifelse(NeuroMET$record_id == "NeuroMet080" & NeuroMET$Visit == "t3" | NeuroMET$record_id == "NeuroMet118" & NeuroMET$Visit == "t2", NA, Sal))

# Calulate years
NeuroMET <- NeuroMET %>%
  group_by(record_id)%>%
  mutate(year = as.numeric(as.Date(visit_date_1, format = "%m/%d/%Y") - 
                             lag(as.Date(visit_date_1, format = "%m/%d/%Y"), default = as.Date(visit_date_1, format = "%m/%d/%Y")[1])
  ))%>%
  mutate(year = cumsum(year))%>%
  mutate(year = round(year/365,2))%>%
  ungroup()%>%
  filter (!is.na(record_id))

NeuroMET <- NeuroMET %>%
  mutate(year2 = case_when(
    year == 0 ~ 0, 
    year >= 0.5 & year < 1.5 ~ 1,
    year >= 1.5 & year < 2.5 ~ 2,
    year >= 2.5 & year < 3.5 ~ 3,
    year >= 3.5 & year < 4.5 ~ 4,
    year >= 4.5 & year < 5.5 ~ 5))


```

# Characteristics of the study sample

```{r table_n_visits, echo=FALSE, warning=FALSE, message=FALSE,  results = "asis"}

table_n  <- NeuroMET %>%
  group_by(diagnose, Visit)%>%
  tally() %>%
  spread(diagnose, n)

table_time <- NeuroMET %>% group_by(diagnose, Visit) %>% summarize("time from baseline" = paste0((round(mean(year), 1)), " (", (round(sd(year), 1)) , ")"))%>% pivot_wider(names_from = diagnose, values_from = "time from baseline") 

row.names_table_1 = c("n", "n", "time from V1 [years]", "n","time from V1 [years]", "n","time from V1 [years]","n","time from V1 [years]")
table_1 <- row.names_table_1 %>% 
  cbind(table_n[1:2,] %>% rbind(table_time[2,]) %>% rbind(table_n[3,])%>% rbind(table_time[3,])%>% rbind(table_n[4,])%>% rbind(table_time[4,])%>% rbind(table_n[5,])%>% rbind(table_time[5,])) %>%
  select(-Visit)

table_1 %>%
  kbl(caption = "**table 1: Observations grouped by diagnosis and visit. Time difference from visit 1 is presented as mean (sd). Further statistical models might contain fewer observations according to data availability.", align = "c")%>%
    kable_styling(bootstrap_options = "condensed", full_width = F)%>%
  pack_rows(group_label = "Visit 1", start_row = 1, end_row = 1) %>%
  pack_rows(group_label = "Visit 2", start_row = 2, end_row = 3) %>%
  pack_rows(group_label = "Visit 3", start_row = 4, end_row = 5) %>%
  pack_rows(group_label = "Visit 4", start_row = 6, end_row = 7) %>%
  pack_rows(group_label = "Visit 5", start_row = 8, end_row = 9)

```


```{r patient_characteristics_v1, results = "asis", out.width = "100%"}
# prepare tests to be used for p-values
ancova_age_edu <- function(data, variable, by, ... ) {
  data <- data[c(variable, by, "age", "edu")] #%>% filter(complete.cases(.)) 
  #  print(data)                              # for check
  lm(data[[variable]] ~ factor(data[[by]]) + data$age + data$edu) %>% anova() %>% broom::tidy() %>% 
    filter (term == "factor(data[[by]])") %>% select(p.value)
}

ancova_age <- function(data, variable, by, ... ) {
  data <- data[c(variable, by, "age")] #%>% filter(complete.cases(.)) 
  lm(data[[variable]] ~ factor(data[[by]]) + data$age) %>% anova() %>% broom::tidy() %>% 
    filter (term == "factor(data[[by]])") %>% select(p.value)
}

ancova_age_weightIns <- function(data, variable, by, ... ) {
  data <- data[c(variable, by, "age", "Ins_CRLB")] #%>% filter(complete.cases(.)) 
  lm(data[[variable]] ~ factor(data[[by]]) + data$age, weights = data$Ins_CRLB) %>% anova() %>% broom::tidy() %>% 
    filter (term == "factor(data[[by]])") %>% select(p.value)
}

ancova_age_weightNAA <- function(data, variable, by, ... ) {
  data <- data[c(variable, by, "age", "NAA_CRLB")] #%>% filter(complete.cases(.)) 
  lm(data[[variable]] ~ factor(data[[by]]) + data$age, weights = data$NAA_CRLB) %>% anova() %>% broom::tidy() %>% 
    filter (term == "factor(data[[by]])") %>% select(p.value)
}

# table consists of 4 stacked tables
theme_gtsummary_compact()

tbl_summary_1 <- 
  NeuroMET%>%
  filter(Visit=="t1") %>%
  select(diagnose, age, gender, edu, APOEe4)%>%
  tbl_summary(
    by = diagnose,
    label = list(age ~ "Age [years]", gender ~ "Female", edu ~ "Education [years]", APOEe4 ~"APOEe4 carrier"),
    missing_text = "N missing",
    digits = list(age ~ 0, edu ~0),
    type = list(c(age, edu) ~ "continuous", c(gender,APOEe4) ~ "dichotomous"),
    value =list (APOEe4 ~ "carrier"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"))%>%
  add_p(pvalue_fun = function(x) style_pvalue(x, digits = 3))%>%
  modify_header(label = "",all_stat_cols() ~ "**{level}**<br>N =  {n}", p.value ~ "**p-value**")%>%
  add_overall(col_label = "**Total**<br> N = {N}")%>%
  modify_spanning_header(c("stat_1","stat_2","stat_3","stat_4") ~ "**Diagnostic Group**")%>%
  modify_caption("**Table 1. Patient's characteristics at visit 1**")%>%
  modify_footnote(c(all_stat_cols(), p.value) ~ NA)


tbl_summary_2 <- 
  NeuroMET%>%
  filter(Visit=="t1") %>%
  select(diagnose, age, edu, mmse, nmm, exefct)%>%
  tbl_summary(
    by = diagnose,
    label = list(mmse ~ "MMSE", nmm ~ "NMM", exefct ~ "Executive function"),
    missing_text = "N missing",
    type = list(c(mmse, nmm, exefct) ~ "continuous"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"))%>%
  add_stat(fns= c("mmse", "nmm", "exefct") ~ ancova_age_edu, 
           pvalue_fun = function(x) style_pvalue(x, digits = 3))%>% 
  modify_header(label = "",all_stat_cols() ~ "**{level}**<br>N =  {n}", p.value ~ "**p-value**")%>%
  add_overall(col_label = "**Total**<br> N = {N}")%>%
  modify_spanning_header(c("stat_1","stat_2","stat_3","stat_4") ~ "**Diagnostic Group**")%>%
  modify_caption("**Table 1. Patient's characteristics at visit 1**")%>%
  modify_footnote(c(all_stat_cols(), p.value) ~ NA)%>%
  remove_row_type(c(age, edu), type = "all")%>%
  modify_fmt_fun(update = p.value ~ function(x) style_pvalue(x, digits = 3))


tbl_summary_3 <- 
  NeuroMET%>%
  filter(Visit=="t1") %>%
  select(diagnose, age, Ins_CRLB, NAA_CRLB, Whole_hip_cs_long, parietal_thick_cs_long, DMN, Sal, Myo, NAA)%>%
  tbl_summary(
    by = diagnose,
    label = list(Whole_hip_cs_long ~ "Hippocampus [mm³]", parietal_thick_cs_long ~ "Parietal Cortical Thickness [mm]" , DMN ~ "DMN connectivity", Sal ~ "Salience connectivity", Myo ~ "Myo-inositol [mmol/L]", NAA ~ "NAA [mmol/L]"),
    missing_text = "N missing",
    type = list(c(Whole_hip_cs_long, DMN, Sal, Myo, NAA) ~ "continuous"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"))%>%
  add_stat(fns= list(c("Whole_hip_cs_long", "parietal_thick_cs_long", "DMN", "Sal") ~ ancova_age, "Myo" ~ ancova_age_weightIns, "NAA" ~ ancova_age_weightNAA), pvalue_fun = function(x) style_pvalue(x, digits = 3))%>% 
  modify_header(label = "", all_stat_cols() ~ "**{level}**<br>N =  {n}", p.value  ="**p-value**")%>%
  add_overall(col_label = "**Total**<br> N = {N}")%>%
  modify_spanning_header(c("stat_1","stat_2","stat_3","stat_4") ~ "**Diagnostic Group**")%>%
  modify_caption("**Table 1. Patient's characteristics at visit 1**")%>%
  modify_footnote(c(all_stat_cols(), p.value) ~ NA)%>%
  remove_row_type(c(age, Ins_CRLB, NAA_CRLB), type = "all")%>%
  modify_fmt_fun(update = p.value ~ function(x) style_pvalue(x, digits = 3))

tbl_summary_4 <- 
  NeuroMET%>%
  filter(Visit=="t1") %>%
  select(diagnose, age, Ab42_40, pTau, GFAP, nfl)%>%
  tbl_summary(
    by = diagnose,
    label = list(Ab42_40 ~ "Ab42/40 [pg/mL]", pTau ~ "p-Tau 181 [pg/mL]", GFAP ~ "GFAP [pg/mL]", nfl ~ "nfl [pg/mL]"),
    missing_text = "N missing",
    type = list(c(Ab42_40, pTau, GFAP, nfl) ~ "continuous"),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"))%>%
  add_stat(fns= list(c("Ab42_40", "pTau", "GFAP", "nfl") ~ ancova_age)) %>%
  modify_header(label = "",all_stat_cols() ~ "**{level}**<br>N =  {n}", p.value ="**p-value**")%>%
  add_overall(col_label = "**Total**<br> N = {N}")%>%
  modify_spanning_header(c("stat_1","stat_2","stat_3","stat_4") ~ "**Diagnostic Group**")%>%
  modify_caption("**Table 2. Patient's characteristics at visit 1**")%>%
  modify_footnote(c(all_stat_cols(), p.value) ~ NA)%>%
  remove_row_type(age, type = "all")%>%
  modify_fmt_fun(update = p.value ~ function(x) style_pvalue(x, digits = 3))

tbl_stack_ex1 <- tbl_stack(list(tbl_summary_1, tbl_summary_2, tbl_summary_3, tbl_summary_4), group_header  = c("Demographics", "Cognition", "MRI/MRS", "Blood-based Biomarkers"))
tbl_stack_ex1 <-tbl_stack_ex1%>%  
  as_gt() %>%
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_row_groups(groups = everything())
  ) %>% 
  gt::tab_source_note(gt::md("Mean (sd) or N(%) are shown for the total study population and for the individual groups. Group differences were tested using Kruskal-Wallis rank sum test (continuous demographics), Pearson's Chi-squared test (dichotomous demographics), multiple regression adjusted for age (MR-based and plasma biomarkers) or adjusted for age and education (cognition). The statistical models for MRS data (myo-inositol and NAA) were additionally weighted based on an estimation of measurement uncertainties of MRS concentrations (Cramér-Rao lower bounds). APOEe4 carrier were identified using TaqMan assays. 
<br>
*Abbreviations: Aß = amyloid beta, AD = Alzheimer’s disease, APOEe4 = apolipoprotein E 4 allele, DMN = default mode network, GFAP = glial fibrillary acidic protein, HC = healthy control, MCI = mild cognitive impairment, MMSE = Mini Mental State Examination (Folstein, Folstein, & McHugh, 1975), NAA = N-Acetylaspartic acid, NfL = neurofilament light chain, NMM = NeuroMET Memory Metric, p-Tau 181 = tau phosphorylated at threonine 181, Sal = Salience network, SCD = subjective cognitive decline.*"))
tbl_stack_ex1

#save table
#tbl_stack_ex1 %>% as_flex_table() %>% flextable::save_as_docx(path = "C:/Users/[...]/Table_2.docx")


```
# Cross-sectional and longitudinal concentrations of plasma biomarkers

```{r Data preparation_long, include = FALSE}
#create a variable diagnose_bl
Diagnose_1 <- NeuroMET %>% filter(Visit == 't1') %>% mutate(diagnose_bl= diagnose) %>% select(record_id, diagnose_bl)
NeuroMET <- NeuroMET %>%
  full_join(Diagnose_1, by = "record_id")

#calculate z-scores of variables (Memory vd Rest and exefct are already z-scaled)
mean_z <- NeuroMET %>%
  filter(diagnose == 'HC' & Visit == 't1') %>%
  summarise(across(c(Ab42_40, pTau, GFAP, nfl, nmm, mmse, Whole_hip_cs_long, parietal_thick_cs_long, DMN, Sal, Myo, NAA), base::mean, na.rm = TRUE))

sd_z <- NeuroMET %>%
  filter(diagnose == 'HC' & Visit == 't1') %>%
  summarise(across(c(Ab42_40, pTau, GFAP, nfl, nmm, mmse, Whole_hip_cs_long, parietal_thick_cs_long, DMN, Sal, Myo, NAA), stats::sd, na.rm = TRUE))

cols <- list("Ab42_40", "pTau", "GFAP", "nfl", "mmse", "nmm", "Whole_hip_cs_long", "parietal_thick_cs_long", "DMN", "Sal", "Myo", "NAA")

NeuroMET[, paste0("z_score_", cols)] <- lapply(cols, function(col){ 
  x <- NeuroMET[col]
  y <- (x - as.vector(mean_z[col])) / as.vector(sd_z[col])
  y[[1]]
  })

#rename
NeuroMET$z_score_hip <- NeuroMET$z_score_Whole_hip_cs_long
NeuroMET$z_score_parietal_thick <- NeuroMET$z_score_parietal_thick_cs_long
NeuroMET$z_score_dmn <- NeuroMET$z_score_DMN
NeuroMET$z_score_sal <- NeuroMET$z_score_Sal
NeuroMET$z_score_myo <- NeuroMET$z_score_Myo
NeuroMET$z_score_naa <- NeuroMET$z_score_NAA

# count available bbm data for each visit
n <- NeuroMET %>% 
  summarise(nAb42_40 = sum(!is.na(Ab42_40)),
            npTau = sum(!is.na(pTau)),
            nGFAP = sum(!is.na(GFAP)),
            nnfl = sum(!is.na(nfl)))

n <- NeuroMET %>%
  group_by(Visit) %>%
  summarise(nAb42_40 = sum(!is.na(Ab42_40)),
            npTau = sum(!is.na(pTau)),
            nGFAP = sum(!is.na(GFAP)),
            nnfl = sum(!is.na(nfl)))%>%
  bind_rows(n)%>%
  ungroup()
```

```{r Models_long_single_variable}
# Define a function for running linear mixed effects models and computing emmeans [[1]] and emtrends [[2]] with their p-values [[3]]
run_lmer_emmeans <- function(response_var, data) {
  data <- data %>% filter(!is.na(.data[[response_var]]) & !is.na(diagnose_bl))
  model <- lmer(paste0(response_var, " ~ diagnose_bl*year + age + (1|record_id)"), 
                data = data, 
                REML = TRUE, na.action = na.omit)
  em <- emmeans(model, pairwise ~diagnose_bl|year, at = list(year = c(0, 1, 3, 5)))
  etrends <- emtrends(model, "diagnose_bl", var= "year") 
  etrendsp <- test(emtrends(model, "diagnose_bl", var= "year"))
  
  return(list(em$emmeans, etrends, etrendsp, em$contrasts, confint(em$contrasts), model))
}
# Alternative for MRS with CRLB weights
run_lmer_emmeans_MRS <- function(response_var, CRLB, data) {
  data <- data %>% filter(!is.na(.data[[response_var]]) & !is.na(diagnose_bl)) %>% mutate(CRLB_2 = .data[[CRLB]])
  #CRLB_2 <- data %>% select(.data[[CRLB]])
  #return(data$CRLB_2)
  model <- lmer(paste0(response_var, " ~ diagnose_bl*year + age + (1|record_id)"),
                data = data,
                weights = CRLB_2,
                REML = TRUE, na.action = na.omit)
  em <- emmeans(model, pairwise ~diagnose_bl|year, at = list(year = c(0, 1, 3, 5)))
  etrends <- emtrends(model, "diagnose_bl", var= "year")
  etrendsp <- test(emtrends(model, "diagnose_bl", var= "year"))
  return(list(em$emmeans, etrends, etrendsp, em$contrasts, confint(em$contrasts), model))
}

# Run the function for each response variable of interest
em_Ab42_40 <- run_lmer_emmeans("Ab42_40", NeuroMET)
em_pTau <- run_lmer_emmeans("pTau", NeuroMET)
em_GFAP <- run_lmer_emmeans("GFAP", NeuroMET)
em_nfl <- run_lmer_emmeans("nfl", NeuroMET)

em_mmse <- run_lmer_emmeans("mmse", NeuroMET)
em_nmm <- run_lmer_emmeans("nmm", NeuroMET)
em_exefct <- run_lmer_emmeans("exefct", NeuroMET)
em_hip <- run_lmer_emmeans("Whole_hip_cs_long", NeuroMET)
em_parietal_thick <- run_lmer_emmeans("parietal_thick_cs_long", NeuroMET)
em_dmn <- run_lmer_emmeans("DMN", NeuroMET)
em_sal <- run_lmer_emmeans("Sal", NeuroMET)
em_myo <- run_lmer_emmeans_MRS(response_var = "Myo", CRLB = "Ins_CRLB", data = NeuroMET)
em_naa <- run_lmer_emmeans_MRS(response_var = "NAA", CRLB = "NAA_CRLB", data = NeuroMET)
```


```{r Figure2, echo=FALSE, warning=FALSE, message=FALSE,  results = "asis", fig.align = "center", fig.height = 10, fig.cap = "**Figure 2: Longitudinal concentration changes of plasma Aß42/Aß40, p-Tau 181, GFAP and NfL.** In the SCD group, only plasma p-Tau 181 showed a relevant yearly increase. The AD group however, showed substantially increasing concentrations of p-Tau 181, GFAP and NfL. Concentrations were measured at visit 1 (n = 127), and when available at follow-ups after 1 year (n = 32), 3 years (n = 44), 4 years (n = 31) and 5 years (n = 9). The thin lines link individual concentrations at each available visit. The linear fits are presented with 95% CI (shaded areas) estimated by linear mixed models adjusted for age. For later interpretation of the models, Orange lines represent the mean of the HC group at visit 1 (continuous line) and 1sd towards the pathological direction (dashed line). *Abbreviations: Aß = amyloid beta, AD = Alzheimer’s disease, CI = confidence interval, GFAP = glial fibrillary acidic protein, HC = healthy control, MCI = mild cognitive impairment, NfL= neurofilament light chain, p-Tau 181 = tau phosphorylated at threonine 181, SCD = subjective cognitive decline.*" }

mean_sd <- c("mean", "1sd") %>% cbind (mean_z %>% rbind(mean_z - sd_z) %>% select(Ab42_40) %>% cbind(mean_z %>% rbind(mean_z + sd_z) %>% select(pTau, GFAP, nfl)))

coeff_Ab42_40 <- as.data.frame(em_Ab42_40[[2]])%>% mutate( 
    coeff= round(year.trend*10^3, 1),
  lower.CL= format(lower.CL*10^3, nsmall = 1, digits = 1), 
  upper.CL= format(upper.CL*10^3, nsmall = 1, digits = 1)) %>% 
  cbind(as.data.frame(em_Ab42_40[[3]]%>% 
                        mutate(p.value = format(round(p.value, 3),nsmall = 3)) %>% 
                        select(p = p.value)))%>%
  mutate(coeff = paste0("ß = ", coeff,  " [", lower.CL, "; ", upper.CL, "]"," * 10<sup>-3</sup>", "<br>p = ", p)) %>% 
  select(diagnose_bl, coeff)
coeff_pTau <- as.data.frame(em_pTau[[2]])%>% mutate(coeff= format(round(year.trend, 2),nsmall = 2), lower.CL= format(round(lower.CL, 2),nsmall = 2), upper.CL= format(round(upper.CL, 2),nsmall = 2)) %>% cbind(as.data.frame(em_pTau[[3]]%>% mutate(p.value = format(round(p.value, 3),nsmall = 3)) %>% select(p = p.value)))%>%
  mutate(coeff = paste0("ß = ", coeff, " [", lower.CL, "; ", upper.CL,"]", '\n', "p = ", p)) %>% select(diagnose_bl, coeff)
coeff_GFAP <- as.data.frame(em_GFAP[[2]])%>% mutate(coeff= format(round(year.trend, 0),nsmall = 0), lower.CL= format(round(lower.CL, 0),nsmall = 0), upper.CL= format(round(upper.CL, 0),nsmall = 0)) %>% cbind(as.data.frame(em_GFAP[[3]]%>% mutate(p.value = format(round(p.value, 3),nsmall = 3)) %>% select(p = p.value)))%>%
  mutate(coeff = paste0("ß = ", coeff, " [", lower.CL, "; ", upper.CL, "]", '\n', "p = ", p)) %>% select(diagnose_bl, coeff)
coeff_nfl <- as.data.frame(em_nfl[[2]])%>% mutate(coeff= format(round(year.trend, 1),nsmall = 1), lower.CL= format(round(lower.CL, 1),nsmall = 1), upper.CL= format(round(upper.CL, 1),nsmall = 1))

p_nfl <- as.data.frame(em_nfl[[3]]%>% mutate(p.value = format(round(p.value, 3),nsmall = 3))) %>% select(p = p.value) 
                                                                                                                                  p_nfl$p[p_nfl$p < 0.001] <- "< 0.001"
p_nfl2 <-c("=", "=", "=", "") %>% cbind(p_nfl)%>% mutate(p = paste0(., " ", p)) %>%select(-".")
coeff_nfl_2<- coeff_nfl%>% cbind(p_nfl2)%>% mutate(coeff = paste0("ß = ", coeff, " [", lower.CL, "; ", upper.CL, "]", '\n', "p ", p)) %>% select(diagnose_bl, coeff)


plot_longitudinal <- function(EM_model, parameter, parameter2, lable_y) {
ggplot (EM_model,aes(x=year, y = emmean, group=diagnose_bl))+
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL, fill = diagnose_bl), alpha = .50) +
  geom_line(aes(colour = diagnose_bl), size = 2)+
  geom_line(data= NeuroMET %>% filter(!is.na(diagnose_bl) & !is.na({{parameter}})), mapping=aes(x = year, y = {{parameter}}, group= record_id, col=diagnose_bl), alpha= 0.5)+
  geom_point(data=NeuroMET %>% filter(!is.na(diagnose_bl) & !is.na({{parameter}})), mapping=aes(x = year, y = {{parameter}}, group= record_id, col=diagnose_bl), size=1, alpha= 0.5)+
  labs(x="Time [years]", y=lable_y, fill = "diagnosis at baseline")+
  facet_wrap(~diagnose_bl,nrow = 1)+
  scale_fill_manual(values =c("gray50", "seagreen", "deepskyblue3" ,"#225EA8"))+
  scale_color_manual(values =c("gray50", "seagreen", "deepskyblue3" ,"#225EA8"))+
  guides(col = "none")+
  theme_bw(base_size=10)+
  theme(legend.position = 'top') 
}

figure_Ab42_40_long <- plot_longitudinal(EM_model = as.data.frame(em_Ab42_40[[1]]), parameter = Ab42_40, parameter2 = "Ab42_40", lable_y= "Aß 42/40 [pg/mL]")+ 
  geom_hline(data= mean_sd %>% filter(. =="mean") , aes(yintercept= Ab42_40), color = "orange", size = 0.4)+    # HC mean
  geom_hline(data= mean_sd %>% filter(. =="1sd"), aes(yintercept= Ab42_40), color = "orange", linetype = "longdash",  size = 0.4)+
  geom_richtext(aes(x=year, y = emmean, label = lab), 
    data = data.frame(year = 5, emmean = 0.1, diagnose_bl = factor(c("HC", "SCD", "MCI", "AD"), levels = c("HC", "SCD", "MCI", "AD")), lab = coeff_Ab42_40$coeff),vjust = "inward", hjust = "inward", size= 3, parse = T,fill = NA, label.colour = NA)+
  scale_y_continuous(breaks = c(0.03, 0.05, 0.07, 0.09, 0.11), labels=c(expression("30*10"^{-3}~""), expression("50*10"^{-3}~""), expression("70*10"^{-3}~""), expression("90*10"^{-3}~""), expression("110*10"^{-3}~"")))


figure_pTau_long <- plot_longitudinal(EM_model = as.data.frame(em_pTau[[1]]), parameter = pTau, parameter2 = "pTau", lable_y= "p-Tau 181 [pg/mL]")+ 
  geom_hline(data= mean_sd %>% filter(. =="mean") , aes(yintercept= pTau), color = "orange", size = 0.4)+    # HC mean
  geom_hline(data= mean_sd %>% filter(. =="1sd"), aes(yintercept= pTau), color = "orange", linetype = "longdash",  size = 0.4)+
  geom_text(aes(x=year, y = emmean, label = lab), 
    data = data.frame(year = 5, emmean = 9, diagnose_bl = factor(c("HC", "SCD", "MCI", "AD"), levels = c("HC", "SCD", "MCI", "AD")), lab = coeff_pTau$coeff),vjust = "inward", hjust = "inward", size= 3)
figure_GFAP_long <- plot_longitudinal(EM_model = as.data.frame(em_GFAP[[1]]), parameter = GFAP, parameter2 = "GFAP", lable_y= "GFAP [pg/mL]")+ 
  geom_hline(data= mean_sd %>% filter(. =="mean") , aes(yintercept= GFAP), color = "orange", size = 0.4)+    # HC mean
  geom_hline(data= mean_sd %>% filter(. =="1sd"), aes(yintercept= GFAP), color = "orange", linetype = "longdash",  size = 0.4)+ 
  geom_text(aes(x=year, y = emmean, label = lab), 
    data = data.frame(year = 5, emmean = 520, diagnose_bl = factor(c("HC", "SCD", "MCI", "AD"), levels = c("HC", "SCD", "MCI", "AD")), lab = coeff_GFAP$coeff),vjust = "inward", hjust = "inward", size= 3)
figure_nfl_long <- plot_longitudinal(EM_model = as.data.frame(em_nfl[[1]]), parameter = nfl, parameter2 = "nfl", lable_y= "NfL [pg/mL]")+ 
  geom_hline(data= mean_sd %>% filter(. =="mean") , aes(yintercept= nfl), color = "orange", size = 0.4)+    # HC mean
  geom_hline(data= mean_sd %>% filter(. =="1sd"), aes(yintercept= nfl), color = "orange", linetype = "longdash",  size = 0.4)+ 
  geom_text(aes(x=year, y = emmean, label = lab), 
    data = data.frame(year = 5, emmean = 130, diagnose_bl = factor(c("HC", "SCD", "MCI", "AD"), levels = c("HC", "SCD", "MCI", "AD")), lab = coeff_nfl_2$coeff),vjust = "inward", hjust = "inward", size= 3)

# arrange plots to one figure using the package patchwork
figure_2 <- (figure_Ab42_40_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())) / 
(figure_pTau_long+ rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank()))/ 
(figure_GFAP_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())) / 
(figure_nfl_long+ theme(strip.background = element_blank(), strip.text.x = element_blank())) +
  plot_layout(guides = 'collect')& theme(legend.position = 'top')
figure_2 

ggsave("figure_2.tiff", height = 11, width = 8, dpi = 600)


```

```{r Figure3, echo=FALSE, warning=FALSE, message=FALSE,  results = "asis", fig.align = "center", fig.height = 4, fig.cap = "**Figure 3: Subgroup analyses of longitudinal trajectories of plasma p-Tau 181, GFAP and NfL based on plasma Aß42/40 changes (decliners / non-decliners).** Aß decliners only showed a substantial increase of p-Tau 181 in the SCD group, but not the HC group. Participants were classified into Aß decliners/non-decliners by a 2-means cluster analysis based on their difference between Aß42/40 at first visit and last visit (mean differenceAß decliner = -0.5sd, nAß  decliner = 22, mean differenceAß non- decliner = 0.2sd, nAß non- decliner = 31). Longitudinal trajectories were extracted from linear mixed models only including participants of the HC and SCD groups, correcting for age, sex and education. *Abbreviations: Aß = amyloid beta, CI = confidence interval, GFAP = glial fibrillary acidic protein, HC = healthy control, NfL= neurofilament light chain, p-Tau 181 = tau phosphorylated at threonine 181, SCD = subjective cognitive decline.*"}

#Create Ab_change (difference between last and first visit)
NeuroMET_Figure3<- NeuroMET %>%
  group_by(record_id)%>%
  filter(row_number() ==1 | row_number() ==n())%>%   #pick first and last visit
  mutate(Ab_diff = z_score_Ab42_40 - lag(z_score_Ab42_40))%>%
  ungroup() 

#create clusters
clusters_2 <- Ckmeans.1d.dp((NeuroMET_Figure3 %>% filter((diagnose_bl =="HC" | diagnose_bl =="SCD") & !is.na(Ab_diff)))$Ab_diff, 2)
#clusters_2

#join clusters to NeuroMET Data
NeuroMET_Figure3_2 <- data.frame(NeuroMET_Figure3 %>% filter((diagnose_bl =="HC" | diagnose_bl =="SCD") & !is.na(Ab_diff)), clusters_2$cluster)%>%
  select(record_id, Ab_diff, clusters_2.cluster)
NeuroMET_Figure3b <- left_join(NeuroMET, NeuroMET_Figure3_2, by = c("record_id"))

#models
long_pTau_abdecliner <- lmer(z_score_pTau ~ year * Ab_diff*diagnose_bl + age + (1 | record_id), 
                               data = NeuroMET_Figure3b %>% filter(diagnose_bl == 'HC'| diagnose_bl == 'SCD'))
em_trends_pTau_abdecliner <-emtrends(long_pTau_abdecliner, pairwise ~ diagnose_bl|Ab_diff, var = "year", at = list(Ab_diff = clusters_2$centers))
em_means_pTau_abdecliner <- emmeans(long_pTau_abdecliner, pairwise ~diagnose_bl|Ab_diff|year, at = list(year = c(0, 5), Ab_diff = clusters_2$centers))

long_GFAP_abdecliner <- lmer(z_score_GFAP ~ year * Ab_diff*diagnose_bl + age + (1 | record_id), 
                             data = NeuroMET_Figure3b %>% filter(diagnose_bl == 'HC'| diagnose_bl == 'SCD'))
em_trends_GFAP_abdecliner <-emtrends(long_GFAP_abdecliner, pairwise ~ diagnose_bl|Ab_diff, var = "year", at = list(Ab_diff = clusters_2$centers))
em_means_GFAP_abdecliner <- emmeans(long_GFAP_abdecliner, pairwise ~diagnose_bl|Ab_diff|year, at = list(year = c(0,5), Ab_diff = clusters_2$centers))

long_nfl_abdecliner <- lmer(z_score_nfl ~ year * Ab_diff*diagnose_bl + age + (1 | record_id), 
                             data = NeuroMET_Figure3b %>% filter(diagnose_bl == 'HC'| diagnose_bl == 'SCD'))
em_trends_nfl_abdecliner <-emtrends(long_nfl_abdecliner, pairwise ~ diagnose_bl|Ab_diff, var = "year", at = list(Ab_diff = clusters_2$centers))
em_means_nfl_abdecliner <- emmeans(long_nfl_abdecliner, pairwise ~diagnose_bl|Ab_diff|year, at = list(year = c(0, 5), Ab_diff = clusters_2$centers))

data_plot <- data.frame(bbm = "p-Tau 181", em_means_pTau_abdecliner$emmeans)%>% mutate(Ab_decliner = factor(Ab_diff, labels = c("Aß decliner", "Aß non-decliner"))) %>%
  rbind(data.frame(bbm = "GFAP", em_means_GFAP_abdecliner$emmeans)%>% mutate(Ab_decliner = factor(Ab_diff, labels = c("Aß decliner", "Aß non-decliner"))))%>%
  rbind(data.frame(bbm = "NfL", em_means_nfl_abdecliner$emmeans)%>% mutate(Ab_decliner = factor(Ab_diff, labels = c("Aß decliner", "Aß non-decliner"))))%>%
  mutate(bbm= as.factor(bbm))

#plots
ggplot (data= data_plot %>% filter(!(bbm=="Aß 42/40")), aes(x = year, y = emmean, color =diagnose_bl))+
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL,fill = diagnose_bl), alpha = .10, colour = NA)+ 
  geom_line(alpha= 0.7, size = 2)+
  ylab("biomarker concentration (z-scaled)") +
  theme_bw(base_size=10)+
  theme(legend.position="top")+
  guides(col = guide_legend(title="Diagnosis at baseline"), fill = "none")+
  scale_x_discrete(limits=c("1", "2","3","4", "5"))+
  scale_fill_manual(values =c("gray50", "seagreen"))+
  scale_color_manual(values =c("gray50", "seagreen"))+
  facet_grid(Ab_decliner~factor(bbm, levels = c('Aß 42/40', 'p-Tau 181', 'GFAP', 'NfL')))
ggsave("figure_3.tiff", height = 4, width = 4, dpi = 600)

#table with all slopes
table_slopes_decliner <- data.frame(em_trends_pTau_abdecliner$emtrends)%>% mutate("p-Tau 181" = paste0(round(year.trend, digits = 2), " [", round(lower.CL,2), "; ", round(upper.CL,2), "]"))%>%
  select (diagnose_bl, Ab_diff,"p-Tau 181") %>% cbind(p = round(data.frame(test(em_trends_pTau_abdecliner)$emtrends)$p.value,3))%>%
  cbind(
    data.frame(em_trends_GFAP_abdecliner$emtrends)%>% mutate("GFAP" = paste0(round(year.trend, digits = 2), " [", round(lower.CL,2), "; ", round(upper.CL,2), "]"))%>%
      select ("GFAP") %>% cbind(p = round(data.frame(test(em_trends_GFAP_abdecliner)$emtrends)$p.value,3))
  )%>%
  cbind(
    data.frame(em_trends_nfl_abdecliner$emtrends)%>% mutate("NfL" = paste0(round(year.trend, digits = 2), " [", round(lower.CL,2), "; ", round(upper.CL,2), "]"))%>%
      select ("NfL") %>% cbind(p = round(data.frame(test(em_trends_nfl_abdecliner)$emtrends)$p.value,3))
  )
#table_slopes_decliner
``` 

# Cross-sectional association between plasma biomarkers and MR-based parameters 

```{r Models_associations_cs_and_long}

#Ab42_40
outcome_vars <- as.list(NeuroMET%>%filter(!is.na(z_score_Ab42_40) & !is.na(diagnose_bl))%>%select(z_score_nmm, exefct, z_score_hip, z_score_parietal_thick, z_score_dmn, z_score_sal))
models_Ab42_40 <- lapply(outcome_vars, function(outcome_var) {
  model_lmer <- lmer(outcome_var ~ z_score_Ab42_40*year + age + gender + edu + (1|record_id),
       data = NeuroMET%>%filter(!is.na(z_score_Ab42_40) & !is.na(diagnose_bl)),
       REML = TRUE, na.action = na.omit)
  emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_Ab42_40", at= list(year = 0)) #crosssectional association
  em_change <- emmeans(model_lmer,pairwise~z_score_Ab42_40*year, at = list(z_score_Ab42_40 = c(0, -1), year = c(0,1)))
  return(list("1", "2", "3", emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
})

#Ab42_40_MRS
models_Ab42_40_MRS <- function(outcome_var, CRLB) {
  data <- NeuroMET %>% mutate(outcome_var_2 = .data[[outcome_var]], CRLB_2= .data[[CRLB]])
  model_lmer <- lmer(outcome_var_2 ~ z_score_Ab42_40*year + age + gender + edu + (1|record_id),
       data = data %>% filter(!is.na(z_score_Ab42_40) & !is.na(diagnose_bl)), 
       weights = CRLB_2, REML = TRUE, na.action = na.omit)
  emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_Ab42_40", at= list(year = 0))
  em_change <- emmeans(model_lmer,pairwise~z_score_Ab42_40*year, at = list(z_score_Ab42_40 = c(0, -1), year = c(0,1)))#crosssectional association
  return(list("1", "2", "3", emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
}

models_Ab42_40_MRS <- mapply(models_Ab42_40_MRS, outcome_var = c("z_score_myo","z_score_NAA"), CRLB = c("Ins_CRLB", "NAA_CRLB"))
models_Ab42_40$z_score_myo <- models_Ab42_40_MRS[1:8]
models_Ab42_40$z_score_NAA <- models_Ab42_40_MRS[9:16]

#pTau
outcome_vars <- as.list(NeuroMET%>%filter(!is.na(z_score_pTau) & !is.na(diagnose_bl))%>%select(z_score_nmm, exefct, z_score_hip, z_score_parietal_thick, z_score_dmn, z_score_sal))
models_pTau <- lapply(outcome_vars, function(outcome_var) {
  model_lmer <- lmer(outcome_var ~ z_score_pTau*year + age + gender + edu + (1|record_id),
       data = NeuroMET%>%filter(!is.na(z_score_pTau) & !is.na(diagnose_bl)),
       REML = TRUE, na.action = na.omit)
  emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_pTau", at= list(year = 0)) #crosssectional association
  em_change <- emmeans(model_lmer,pairwise~z_score_pTau*year, at = list(z_score_pTau = c(0, 1), year = c(0,1)))
  return(list("1", "2", "3",emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
})
#pTau_MRS
models_pTau_MRS <- function(outcome_var, CRLB, data) {
  data <- NeuroMET %>% mutate(outcome_var_2 = .data[[outcome_var]], CRLB_2= .data[[CRLB]])
  model_lmer <- lmer(outcome_var_2 ~ z_score_pTau*year + age + gender + edu + (1|record_id),
       data = data %>% filter(!is.na(z_score_pTau) & !is.na(diagnose_bl)), 
       weights = CRLB_2, REML = TRUE, na.action = na.omit)
  emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_pTau", at= list(year = 0)) #crosssectional association
  em_change <- emmeans(model_lmer,pairwise~z_score_pTau*year, at = list(z_score_pTau = c(0, 1), year = c(0,1)))
  return(list("1", "2", "3",emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
}
models_pTau_MRS <- mapply(models_pTau_MRS, outcome_var = c("z_score_myo","z_score_NAA"), CRLB = c("Ins_CRLB", "NAA_CRLB"))
models_pTau$z_score_myo <- models_pTau_MRS[1:8]
models_pTau$z_score_NAA <- models_pTau_MRS[9:16]
#GFAP
outcome_vars <- as.list(NeuroMET%>%filter(!is.na(z_score_GFAP) & !is.na(diagnose_bl))%>%select(z_score_nmm, exefct, z_score_hip, z_score_parietal_thick, z_score_dmn, z_score_sal))
models_GFAP <- lapply(outcome_vars, function(outcome_var) {
  model_lmer <- lmer(outcome_var ~ z_score_GFAP*year + age + gender + edu + (1|record_id),
       data = NeuroMET%>%filter(!is.na(z_score_GFAP) & !is.na(diagnose_bl)),
       REML = TRUE, na.action = na.omit)
  emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_GFAP", at= list(year = 0)) #crossectional association
  em_change <- emmeans(model_lmer,pairwise~z_score_GFAP*year, at = list(z_score_GFAP = c(0, 1), year = c(0,1)))
 return(list("1", "2", "3",emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
})
#GFAP_MRS
models_GFAP_MRS <- function(outcome_var, CRLB, data) {
  data <- NeuroMET %>% mutate(outcome_var_2 = .data[[outcome_var]], CRLB_2= .data[[CRLB]])
  model_lmer <- lmer(outcome_var_2 ~ z_score_GFAP*year + age + gender + edu + (1|record_id),
       data = data %>% filter(!is.na(z_score_GFAP) & !is.na(diagnose_bl)), 
       weights = CRLB_2, REML = TRUE, na.action = na.omit)
 emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_GFAP", at= list(year = 0)) #crossectional association
  em_change <- emmeans(model_lmer,pairwise~z_score_GFAP*year, at = list(z_score_GFAP = c(0, 1), year = c(0,1)))
 return(list("1", "2", "3",emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
}
models_GFAP_MRS <- mapply(models_GFAP_MRS, outcome_var = c("z_score_myo","z_score_NAA"), CRLB = c("Ins_CRLB", "NAA_CRLB"))
models_GFAP$z_score_myo <- models_GFAP_MRS[1:8]
models_GFAP$z_score_NAA <- models_GFAP_MRS[9:16]
#nfl
outcome_vars <- as.list(NeuroMET%>%filter(!is.na(z_score_nfl) & !is.na(diagnose_bl))%>%select(z_score_nmm, exefct, z_score_hip, z_score_parietal_thick, z_score_dmn, z_score_sal))
models_nfl <- lapply(outcome_vars, function(outcome_var) {
  model_lmer <- lmer(outcome_var ~ z_score_nfl*year + age + gender + edu + (1|record_id),
       data = NeuroMET%>%filter(!is.na(z_score_nfl) & !is.na(diagnose_bl)),
       REML = TRUE, na.action = na.omit)
  emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_nfl", at= list(year = 0)) #crossectional association
  em_change <- emmeans(model_lmer,pairwise~z_score_nfl*year, at = list(z_score_nfl = c(0, 1), year = c(0,1)))
 return(list("1", "2", "3",emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
})

#nfl_MRS
models_nfl_MRS <- function(outcome_var, CRLB, data) {
  data <- NeuroMET %>% mutate(outcome_var_2 = .data[[outcome_var]], CRLB_2= .data[[CRLB]])
  model_lmer <- lmer(outcome_var_2 ~ z_score_nfl*year + age + gender + edu + (1|record_id),
       data = data %>% filter(!is.na(z_score_nfl) & !is.na(diagnose_bl)), 
       weights = CRLB_2, REML = TRUE, na.action = na.omit)
  emtrends_cs <- emtrends(model_lmer, "year", var = "z_score_nfl", at= list(year = 0)) #crossectional association
  em_change <- emmeans(model_lmer,pairwise~z_score_nfl*year, at = list(z_score_nfl = c(0, 1), year = c(0,1)))
  return(list("1", "2", "3",emtrends_cs, nobs(model_lmer), nobs_bl = (summary(model_lmer))[3]$devcomp$dims[5], model_lmer, em_change))
}
models_nfl_MRS <- mapply(models_nfl_MRS, outcome_var = c("z_score_myo","z_score_NAA"), CRLB = c("Ins_CRLB", "NAA_CRLB"))
models_nfl$z_score_myo <- models_nfl_MRS[1:8]
models_nfl$z_score_NAA <- models_nfl_MRS[9:16]
```

```{r Figure_4, echo=FALSE, warning=FALSE, message=FALSE, fig.height = 7, fig.width = 5, results = "asis", fig.align = "center", fig.cap = "**Figure 4: Associations between concentrations of plasma Aß42/40, p-Tau 181, GFAP and NfL and z-transformed parameters of cognition, structural and functional MRI, and MRS.** Largest effect sizes were consistently found for either p-Tau 181 or GFAP. Effects for Aß42/40 were smaller with large uncertainties. Effects for NMM, i.e. memory ability, were considerably larger than for any of the MR-based parameters. Depending on data availability, observations of a maximum of 127 participants were included. The effects are presented with 95% CI (horizontal bars) estimated by linear mixed models adjusted for age, sex and education. Models including the MRS parameters Myo-inositol and NAA were additionally weighted based on a measure of uncertainty (CRLB). Connectivity measures for the DMN and Sal were acquired at resting-state. *Abbreviations: Aß = amyloid beta, CI = confidence interval, CRLB = Cramer-Rao Lower Bound, DMN = Default mode network, GFAP = glial fibrillary acidic protein, NAA = N-Acetylaspartic acid, NfL= neurofilament light chain, NMM = NeuroMET memory metric, obs. = observations, part. = participants, p-Tau 181 = tau phosphorylated at threonine 181, Sal = Salience network. *" }

coefficients_figure <- function (model) {
  model_emtrends <- model[[4]] 
  c(n = as.data.frame(model[[6]]), trend = as.data.frame(model_emtrends)[1,2], lower = as.data.frame(model_emtrends)[1,5], upper = as.data.frame(model_emtrends)[1,6], p = as.data.frame(test(model_emtrends))[1,6])
}

m_Ab42_40 <- mapply(coefficients_figure, models_Ab42_40[1:8])
m_pTau <- mapply(coefficients_figure, models_pTau)
m_GFAP <- mapply(coefficients_figure, models_GFAP)
m_nfl <- mapply(coefficients_figure, models_nfl)
df <- data.frame(BBM = "Ab42_40", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_Ab42_40)))%>%
  rbind (data.frame(BBM = "pTau", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_pTau))))%>%
  rbind (data.frame(BBM = "GFAP", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_GFAP))))%>%
  rbind (data.frame(BBM = "nfl", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_nfl)))) %>%
  mutate(BBM = factor(BBM, levels = c("Ab42_40", "pTau", "GFAP", "nfl"), labels= c("Aß 42/40", "p-Tau 181", "GFAP", "NfL")))
row.names(df) <- NULL

df_Figure <- df %>% 
  mutate(
  outcome_vars = outcome_vars %>%
    fct_recode("NMM" = "z_score_nmm", "Executive Function"  = "exefct", "Hippocampus Volume" = "z_score_hip", "Parietal Cortical Thickness" = "z_score_parietal_thick", "DMN Connectivity" = "z_score_dmn","Sal Connectivity" = "z_score_sal",  "Myo-inositol" = "z_score_myo","NAA" = "z_score_NAA")%>% fct_relevel("NMM" , "Executive Function" , "Hippocampus Volume", "Parietal Cortical Thickness" ,"DMN Connectivity","Sal Connectivity", "Myo-inositol" ,"NAA")
  )%>%
  mutate(BBM = factor(BBM, levels = c("NfL", "GFAP", "p-Tau 181", "Aß 42/40")))

Figure4 <- df_Figure %>%
  ggplot(mapping = aes(x = as.vector(unlist(trend)),
                     y = outcome_vars,
                     color = BBM))+
  geom_point(position = position_dodge(width =0.5), size = 1)+
  geom_linerange(aes(xmin=as.vector(unlist(lower)), xmax=as.vector(unlist(upper))), position = position_dodge(width =0.5), size= 0.8)+
  xlab(label = "std. ß coefficient")+
  ylab(NULL)+
  guides(color = guide_legend(title = "Plasma biomarkers", reverse= TRUE))+
  ylim(rev(levels(df_Figure$outcome_vars)))+ # falls die reihenfolge der Outcome_vars nicht stimmt
  geom_vline(xintercept = 0, color = "dark grey")+
  theme_minimal()+
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF"),
        legend.position = "right",
        legend.text=element_text(size=12),
        legend.key.size =  unit(0.5, 'cm'),
        legend.title = element_text(size  = 12),
        panel.spacing = unit(1.5, "lines"),
        axis.text.y= element_blank(),
        axis.text.x=element_text(size=10),
        #axis.title.x = element_text(size=10)
        )+
  scale_color_manual(values=c('tomato', 'steelblue', 'seagreen', 'orange2'))

#Add nobs in figure:
figure_outcome_vars <- data.frame(outcome_vars_figure = c("NMM", "Executive Function", "Hippocampus Volume", "Parietal Cortical Thickness","DMN Connectivity", "Sal Connectivity", "Myo-inositol", "NAA")) %>% cbind(nparticipants = unlist(df_Figure$n.model..6..[1:8]))%>%
  mutate(outcome_vars_figure = factor(outcome_vars_figure, levels = rev(outcome_vars_figure)))

p_names <- figure_outcome_vars %>%
  ggplot(aes(y = factor(outcome_vars_figure)))+
  geom_text(aes(x = 0, label = outcome_vars_figure), hjust = "right", size = 4)+
  theme_void()

n <- figure_outcome_vars %>%
  ggplot () +
geom_text(
    aes(x = 0, 
        y = outcome_vars_figure, label = nparticipants),
    hjust = 0, size = 4,
    fontface = "plain")+
  theme_void() +
  ggtitle("n")+ theme(plot.title = element_text(size = 11, face = "bold", hjust =0.6, vjust = 0))

layout <- c(
  area(t = 0, l = 0, b = 15, r = 60), 
  area(t = 0, l = 25, b = 15, r = 45), 
  area(t = 0, l = 47, b = 15, r = 70)
  )
# final plot arrangement
Figure_4b <- p_names + n + Figure4 + plot_layout(design = layout)
Figure_4b
ggsave("Figure_4.tiff", width = 7.5, height = 6, dpi = 600)
```

```{r table_A2, echo=FALSE, warning=FALSE, message=FALSE, results = "asis", fig.align = "center" }

table_associations_cs_1 <- df %>%
  mutate("std. ß[95% CI]" = paste0(round(as.vector(unlist(trend)),2), " [", round(as.vector(unlist(lower)),2), "; ", round(as.vector(unlist(upper)),2), "]"),
         "p" = round(as.vector(unlist(p)),3))%>%
  select(BBM, outcome_vars, n = as.vector(unlist("n.model..6..")), "std. ß[95% CI]", p)%>%
  mutate(outcome_vars = outcome_vars %>%
    fct_recode("NMM" = "z_score_nmm", "Executive Function"  = "exefct", "Hippocampus Volume" = "z_score_hip", "Parietal Cortical Thickness" = "z_score_parietal_thick", "DMN Connectivity" = "z_score_dmn","Sal Connectivity"= "z_score_sal",  "Myo-inositol" = "z_score_myo","NAA" = "z_score_NAA")%>% fct_relevel(
     "NMM" , "Executive Function" , "Hippocampus Volume", "Parietal Cortical Thickness" ,"DMN Connectivity", "Sal Connectivity",  "Myo-inositol" ,"NAA"))

table_associations_cs_1$p[table_associations_cs_1$p <  0.001] <- "<0.001"
table_associations_cs_1 <- table_associations_cs_1%>% mutate(p = ifelse(p <0.05, paste0("**", p, "**"),p))
table_associations_cs_1<- table_associations_cs_1%>%arrange(outcome_vars)

table_associations_cs_2 <- table_associations_cs_1%>%filter(BBM =="Aß 42/40")%>% select (outcome_vars, n, "std. ß[95% CI]", p) %>%
  cbind(table_associations_cs_1%>%filter(BBM =="p-Tau 181") %>% select ("std. ß[95% CI]", p))%>%
  cbind(table_associations_cs_1%>%filter(BBM =="GFAP")%>% select ("std. ß[95% CI]", p))%>%
  cbind(table_associations_cs_1%>%filter(BBM =="NfL")%>% select ("std. ß[95% CI]", p))
options(knitr.kable.NA = "")

table_A2 <- kable (table_associations_cs_2,  col.names = c("z-scores", "n", "std. ß [95% CI]", "p", "std. ß [95% CI]", "p", "std. ß [95% CI]", "p", "std. ß [95% CI]", "p"), caption = "**Table A2: Associations between concentrations of plasma Aß42/40, p-Tau 181, GFAP and NfL and parameters of cognition, structural and functional MRI and MRS at visit 1.**", align = "c", format="html") %>%
  kable_classic() %>%
  add_header_above(c("  " = 2, "**Aß 42/40**" = 2, "**p-Tau 181**" = 2, "**GFAP**" = 2, "**Nfl**" = 2))%>%
  footnote("The effects [95% CI] were estimated by linear mixed models adjusted for age, sex and education. Models including the MRS markers Myo-inositol and NAA were additionally weighted based on a measure of uncertainty (CRLB). All plasma biomarkers and outcome variables were z-standardized in order to allow for comparisons.*Abbreviations: Aß = amyloid beta, CI = confidence interval, CRLB = Cramer-Rao Lower Bound, DMN = default mode network, GFAP = glial fibrillary acidic protein, NAA = N-Acetylaspartic acid, NfL= neurofilament light chain, NMM = NeuroMET Memory Metric, p-Tau 181 = phosphorylated tau at threonine 181, Sal = Salience network.*",
           general_title = "")

table_A2
```

# Longitudinal association between bbms and outcome measures

```{r changes_long, echo=FALSE, warning=FALSE, message=FALSE, results = "asis", fig.align = "center" }
coefficients_changes_long <- function (model) {
  em_changes <- model[[8]] 
  as.data.frame(c(n_obs = as.data.frame(model[[5]]), n_participants = as.data.frame(model[[6]]), 
                  change0_0 = data.frame(em_changes$contrasts[2])$estimate, 
                  change0_0_lower = (confint(em_changes)$contrasts)$lower.CL[2], 
                  change0_0_upper = (confint(em_changes)$contrasts)$upper.CL[2], 
                  change0_0_p = data.frame(unlist(em_changes)$contrasts)$`p.value`[2], 
                  change0_1 = data.frame(em_changes$contrasts[3])$estimate, 
                  change0_1_lower = (confint(em_changes)$contrasts)$lower.CL[3], 
                  change0_1_upper = (confint(em_changes)$contrasts)$upper.CL[3], 
                  change0_1_p = data.frame(unlist(em_changes)$contrasts)$`p.value`[3],
                  change1_1 = data.frame(em_changes$contrasts[5])$estimate, 
                  change1_1_lower = (confint(em_changes)$contrasts)$lower.CL[5], 
                  change1_1_upper = (confint(em_changes)$contrasts)$upper.CL[5], 
                  change1_1_p = data.frame(unlist(em_changes)$contrasts)$`p.value`[5]
    ))
}

m_Ab42_40 <- mapply(coefficients_changes_long, models_Ab42_40)  
m_pTau <- mapply(coefficients_changes_long, models_pTau)
m_GFAP <- mapply(coefficients_changes_long, models_GFAP)
m_nfl <- mapply(coefficients_changes_long, models_nfl)
df <- data.frame(bbm = "Ab42_40", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_Ab42_40)))%>% 
  rbind (data.frame(bbm = "pTau", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_pTau))))%>%
  rbind (data.frame(bbm = "GFAP", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_GFAP))))%>%
  rbind (data.frame(bbm = "nfl", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_nfl))))%>%
  mutate(bbm = factor(bbm, levels = c("Ab42_40", "pTau", "GFAP", "nfl"), labels= c("Aß 42/40", "p-Tau 181", "GFAP", "NfL")))
row.names(df) <- NULL
```

```{r Figure_5, echo=FALSE, warning=FALSE, message=FALSE,  results = "asis", out.width = "100%", fig.align = "center", fig.cap="**Figure 5: Yearly change of cognitive, structural and functional MRI, and MRS parameters depending on changes of plasma Aß42/40, p-Tau 181, GFAP and NfL.** To allow for comparison, all parameters were z-transformed. Yearly changes were estimated for discrete changes of plasma biomarker concentrations after one year: stable concentrations of the mean of the HC group (normal stable, blue), change of concentration from the mean of the HC group to 1sd towards the pathological direction as indicated in figure 1 (normal > abnormal, red), stable concentrations of 1sd towards the pathological direction (abnormal stable, green). Depending on data availability, a maximum of 127 participants with 243 observations were included. The effects are presented with 95% CI estimated by linear mixed models adjusted for age, sex and education. Models including the MRS markers Myo-inositol and NAA were additionally weighted based on a measure of uncertainty (CRLB). *Abbreviations: Aß = amyloid beta, CI = confidence interval, CRLB = Cramer-Rao Lower Bound, DMN = Default mode network, GFAP = glial fibrillary acidic protein, HC = healthy control, NAA = N-Acetylaspartic acid, NfL= neurofilament light chain, NMM = NeuroMET Memory Metric, p-Tau 181 = tau phosphorylated at threonine 181, Sal = Salience network.*"}

names_change <- c(paste0("normal stable"),paste0("normal > abnormal"), paste0("abnormal stable"))
names_change <- rep(names_change, 32 )

df_Figure5 <- df %>%
        select(bbm, outcome_vars, n_obs.model..5.., n_participants.model..6.., change0_0, change0_1, change1_1) %>%
        pivot_longer(cols = -c(1,2,3,4), names_to = "bbm_change", values_to = "z_Score") %>%
  cbind(df %>% select(bbm, outcome_vars, n_obs.model..5.., n_participants.model..6.., ends_with("lower")) %>% 
        pivot_longer(cols = -c(1,2,3,4), names_to = "bbm_change", values_to = "lowerCI")%>%
        select (lowerCI))%>%
  cbind(df %>% select(bbm, outcome_vars, n_obs.model..5.., n_participants.model..6.., ends_with("upper")) %>% 
        pivot_longer(cols = -c(1,2,3,4), names_to = "bbm_change", values_to = "upperCI")%>%
        select (upperCI))%>% 
  mutate(
   outcome_vars = (outcome_vars %>%
    fct_recode("NMM" = "z_score_nmm", "Executive Function"  = "exefct", "Hippocampus Volume" = "z_score_hip", "Parietal Cortical Thickness" = "z_score_parietal_thick", "DMN Connectivity" = "z_score_dmn", "Sal Connectivity"= "z_score_sal",  "Myo-inositol" = "z_score_myo","NAA" = "z_score_NAA")%>% fct_relevel("NMM", "Executive Function" , "Hippocampus Volume", "Parietal Cortical Thickness" ,"DMN Connectivity","Sal Connectivity", "Myo-inositol" ,"NAA")))%>%
    cbind(names_change) %>%
    mutate(names_change = factor(names_change, levels = c("normal > abnormal", "abnormal stable", "normal stable"))) 

Figure5 <- df_Figure5 %>%
  unnest()%>%
    mutate_if(is.numeric, funs(. * -1)) %>%
  ggplot(mapping = aes(x = z_Score,
                     y = outcome_vars,
                     color = as.factor(names_change)))+
  geom_point(position = position_dodge(width =0.5), size = 0.8)+
  geom_linerange(aes(xmin=lowerCI, xmax=upperCI), position = position_dodge(width =0.5), size= 0.7)+
  xlab(label = "yearly change (z-score)")+
  ylab(NULL)+
  guides(color = guide_legend(title = "Plasma biomarker change after 1 year",reverse = TRUE, title.position = "top", title.hjust = 0.5))+
  facet_grid(~bbm)+
  ylim(rev(levels(df_Figure5$outcome_vars)))+ 
  geom_vline(xintercept = 0, color = "dark grey")+
  theme_minimal()+
  theme(strip.placement = "outside",
        strip.background = element_rect(fill = "#EEEEEE", color = "#FFFFFF"),
        legend.position = "top",
        legend.text=element_text(size=9),
        legend.key.size =  unit(0.5, 'cm'),
        legend.title = element_text(size  = 9),
        axis.text.y= element_blank(),
        panel.spacing = unit(1.5, "lines"))+
  scale_color_manual(values=c('orange2', 'steelblue','tomato','seagreen'))
Figure5

figure_outcome_vars <- data.frame(outcome_vars_figure = c("NMM", "Executive Function", "Hippocampus Volume", "Parietal Cortical Thickness","DMN Connectivity", "Sal Connectivity", "Myo-inositol", "NAA")) %>% cbind(nparticipants = unlist((df_Figure5%>%filter(bbm_change == "change0_0") %>% select(n_participants.model..6..)))[1:8])%>%
cbind(nobs = unlist((df_Figure5%>%filter(bbm_change == "change0_0") %>% select(n_obs.model..5..)))[1:8])%>%
  mutate(outcome_vars_figure = factor(outcome_vars_figure, levels = rev(outcome_vars_figure)))

p_names <- figure_outcome_vars %>%
  #mutate(outcome_vars_figure = factor(outcome_vars_figure, levels = rev(outcome_vars_figure)))
  ggplot(aes(y = factor(outcome_vars_figure)))+
  geom_text(aes(x = 0, label = outcome_vars_figure), hjust = "right", size = 3.2,
    color = "gray20")+
  theme_void()

npart <- figure_outcome_vars %>%
  ggplot () +
geom_text(
    aes(x = 0, 
        y = outcome_vars_figure, label = nparticipants),
    hjust = 0.5, size = 3.2,
    fontface = "plain",
    color = "gray20")+
  theme_void() +
  ggtitle("n \n part.")+ theme(plot.title = element_text(size = 9, face = "bold", hjust =0.5, vjust = -33, color = "gray30"))

nobs <- figure_outcome_vars %>%
  ggplot () +
geom_text(
    aes(x = 0, 
        y = outcome_vars_figure, label = nobs),
    hjust = 0.5, size = 3.2,
    fontface = "plain",
    color = "gray20")+
  theme_void() +
  ggtitle("n \n obs.")+ theme(plot.title = element_text(size = 9, face = "bold", hjust =0.5, vjust = -33, color = "gray30"))
  

layout <- c(
  area(t = 0, l = 0, b = 15, r = 110), 
  area(t = 0, l = 58, b = 15, r = 73),
  area(t = 0, l = 73, b = 15, r = 83),
  area(t = 0, l = 85, b = 15, r = 200)
  )
# final plot arrangement
Figure_5b <- p_names + npart + nobs + Figure5 + plot_layout(design = layout)
Figure_5b
ggsave("Figure_5b.tiff", dpi = 600)

```

# Supplemental Material 

```{r figureA1, results = "asis", fig.height = 3, out.width = "100%", fig.cap = "**Figure A 1: Concentration of plasma biomarkers at visit 1.** Horizontals lines indicate the mean of the HC group (continuous line) and 1sd towards the pathological direction (long dashed line). **Abbreviations: AD = Alzheimer’s disease, Aß = amyloid beta, GFAP = glial fibrillary acidic protein, HC = healthy control, MCI = mild cognitive impairment, NfL= neurofilament light chain, p-Tau 181 = tau phosphorylated at threonine 181, SCD = subjective cognitive decline.**"}

breaks_mean <- mean_sd%>%pivot_longer (cols = -c(1), names_to = "bbm", values_to = "mean")%>%filter(. == "mean")%>%mutate(bbm= c('Aß 42/40', 'p-Tau 181', 'GFAP','Nfl'))%>% select(-.)
breaks_sd <- mean_sd%>%pivot_longer (cols = -c(1), names_to = "bbm", values_to = "sd")%>%filter(. == "1sd")%>%mutate(bbm= c('Aß 42/40', 'p-Tau 181', 'GFAP','Nfl'))%>% select(-.)

figure_A1 <- NeuroMET %>%
  filter(Visit == "t1")%>%
  select(record_id, diagnose, "Aß 42/40" = Ab42_40, "p-Tau 181" = pTau, "GFAP" = GFAP, "Nfl" = nfl)%>%
  pivot_longer(cols = 3:6, names_to = "bbm", values_drop_na = TRUE)%>%
  group_by(diagnose)%>%
  ggplot()+ 
  aes(x= diagnose, y= value, fill = diagnose)+
  geom_boxplot(outlier.shape = NA, alpha = 0.95, lwd = 0.3, alpha = 0.6)+
  geom_jitter(position=position_jitter(0.2), size = 1, alpha = 0.4)+
  xlab(NULL)+
  ylab("concentration pg/mL") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))+ ####remove background grid
  scale_fill_manual(values=c("#F8766D", "#00BFC4" , "#CD9600","gray50"))+
  facet_wrap(~factor(bbm, levels = c('Aß 42/40', 'p-Tau 181', 'GFAP','Nfl')), scales = "free", ncol = 4)+
  geom_hline(data= breaks_mean, aes(yintercept= mean), color = "dark grey", size = 0.3)+
  geom_hline(data= breaks_sd, aes(yintercept= sd), linetype = "longdash", color = "dark grey", size = 0.3)+
  theme(legend.position = "none")
figure_A1
ggsave("figure_A1.tiff", height = 3,  dpi = 600)

```



```{r FigureA2, echo=FALSE, warning=FALSE, message=FALSE,  results = "asis", fig.align = "center", fig.height = 24, fig.cap = "**Figure A2: Longitudinal changes of cognitive and MR-based parameters.** The thin lines link individual concentrations at each available visit. The linear fits are presented with 95% CI (shaded areas) estimated by linear mixed models adjusted for age (analogue to model 2). Models including MRS measurements were weighted based on an estimate for measurement uncertainty (CRLB). *Abbreviations: AD = Alzheimer’s disease, CI = confidence interval, CRLB = Cramér-Rao lower bound, HC = healthy control, MCI = mild cognitive impairment, NAA = N-Acetylaspartic acid, NMM= NeuroMET Memory Metric, SCD = subjective cognitive decline.* "}

figure_nmm_long <- plot_longitudinal(EM_model = as.data.frame(em_nmm[[1]]), parameter = nmm, parameter2 = "nmm", lable_y= "NMM")
figure_exefct_long <- plot_longitudinal(EM_model = as.data.frame(em_exefct[[1]]), parameter = exefct, parameter2 = "exefct", lable_y= "Executive function")
figure_hip_long <- plot_longitudinal(EM_model = as.data.frame(em_hip[[1]]), parameter = Whole_hip_cs_long, parameter2 = "Whole_hip_cs_long", lable_y= "Hippocampus volume [mm³]")
figure_parietal_long <- plot_longitudinal(EM_model = as.data.frame(em_parietal_thick[[1]]), parameter = parietal_thick_cs_long, parameter2 = "parietal_thick_cs_long", lable_y= "Parietal thickness [mm²]")
figure_dmn_long <- plot_longitudinal(EM_model = as.data.frame(em_dmn[[1]]), parameter = DMN, parameter2 = "DMN", lable_y= "Default Mode Network")
figure_sal_long <- plot_longitudinal(EM_model = as.data.frame(em_sal[[1]]), parameter = Sal, parameter2 = "Sal", lable_y= "Salience Network")
figure_myo_long <- plot_longitudinal(EM_model = as.data.frame(em_myo[[1]]), parameter = Myo, parameter2 = "Myo", lable_y= "Myoinositol [mmol/L]")
figure_naa_long <- plot_longitudinal(EM_model = as.data.frame(em_naa[[1]]), parameter = NAA, parameter2 = "NAA", lable_y= "NAA [mmol/L]")

# arrange plots to one figure using the package patchwork
figure_A2 <- (figure_nmm_long+ rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()))/ 
(figure_exefct_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())) /
(figure_hip_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())) / 
(figure_parietal_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())) /
(figure_dmn_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())) / 
(figure_sal_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())) /
(figure_myo_long + rremove("xlab") + theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())) /
  (figure_naa_long + theme(strip.background = element_blank(), strip.text.x = element_blank()))/
plot_layout(guides = 'collect')& theme(legend.position = 'bottom')
figure_A2
ggsave("figure_A2.tiff", height = 14, width = 8, dpi = 600)

```

```{r Table_A3, echo=FALSE, warning=FALSE, message=FALSE, results = "asis", fig.align = "center" }
coefficients_long <- function (model) {
  model_lmer <- model[[7]] 
  confint_lmer <- confint(model_lmer)
  as.data.frame(c(n_obs = as.data.frame(model[[5]]), n_participants = as.data.frame(model[[6]]), fixef_interact = fixef(model_lmer)[7], fixef_interact_lower = confint_lmer[9,1], fixef_interact_upper = confint_lmer[9,2], p = (car::Anova(model_lmer)$`Pr(>Chisq)`)[6]  
    ))
}

m_Ab42_40 <- mapply(coefficients_long, models_Ab42_40)  
m_pTau <- mapply(coefficients_long, models_pTau)
m_GFAP <- mapply(coefficients_long, models_GFAP)
m_nfl <- mapply(coefficients_long, models_nfl)
df <- data.frame(bbm = "Ab42_40", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_Ab42_40))%>% select("n obs"= n_obs.model..5.., "n participants" = n_participants.model..6.., "interaction" = fixef_interact.z_score_Ab42_40.year, "lower" =fixef_interact_lower, "upper" = fixef_interact_upper, p))%>%
  rbind (data.frame(bbm = "pTau", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_pTau))%>% select("n obs"= n_obs.model..5.., "n participants" = n_participants.model..6.., "interaction" = fixef_interact.z_score_pTau.year, "lower" =fixef_interact_lower, "upper" = fixef_interact_upper, p)))%>%
  rbind (data.frame(bbm = "GFAP", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_GFAP))%>% select("n obs"= n_obs.model..5.., "n participants" = n_participants.model..6.., "interaction" = fixef_interact.z_score_GFAP.year, "lower" =fixef_interact_lower, "upper" = fixef_interact_upper, p)))%>%
  rbind (data.frame(bbm = "nfl", outcome_vars = c(names(outcome_vars), "z_score_myo","z_score_NAA"), as.data.frame(t(m_nfl))%>% select("n obs"= n_obs.model..5.., "n participants" = n_participants.model..6.., "interaction" = fixef_interact.z_score_nfl.year, "lower" =fixef_interact_lower, "upper" = fixef_interact_upper, p))) %>%
  mutate(bbm = factor(bbm, levels = c("Ab42_40", "pTau", "GFAP", "nfl"), labels= c("Ab 42/40", "p-Tau 181", "GFAP", "NfL")))
row.names(df) <- NULL
table_associations_long_1 <- df %>%
  mutate("std. ß[95% CI]" = paste0(round(as.vector(unlist(interaction)),2), " [", round(as.vector(unlist(lower)),2), "; ", round(as.vector(unlist(upper)),2), "]"),
         "p" = round(as.vector(unlist(p)),3))%>%
  select(bbm, outcome_vars, "n participants" = as.vector(unlist("n.participants")), "n obs" = as.vector(unlist("n.obs")), "std. ß[95% CI]", p)%>%
  mutate(outcome_vars = outcome_vars %>%
    fct_recode("NMM" = "z_score_nmm", "Executive Function"  = "exefct", "Hippocampus Volume" = "z_score_hip", "Parietal Cortical Thickness" = "z_score_parietal_thick", "DMN Connectivity" = "z_score_dmn", "Sal Connectivity" = "z_score_sal",  "Myo-inositol" = "z_score_myo","NAA" = "z_score_NAA")%>% fct_relevel( "NMM", "Executive Function" , "Hippocampus Volume", "Parietal Cortical Thickness" ,"DMN Connectivity", "Sal Connectivity", "Myo-inositol" ,"NAA"))

table_associations_long_1$p[table_associations_long_1$p <  0.001] <- "<0.001"
table_A3a <- table_associations_long_1%>% mutate(p = ifelse(p <0.05, paste0("**", p, "**"),p))
table_A3a<- table_associations_long_1%>%arrange(outcome_vars)
table_associations_long_2 <- table_associations_long_1%>%filter(bbm =="Ab 42/40")%>% select (outcome_vars, "n obs", "n participants", "std. ß[95% CI]", p) %>%
  cbind(table_associations_long_1%>%filter(bbm =="p-Tau 181") %>% select ("std. ß[95% CI]", p))%>%
  cbind(table_associations_long_1%>%filter(bbm =="GFAP")%>% select ("std. ß[95% CI]", p))%>%
  cbind(table_associations_long_1%>%filter(bbm =="NfL")%>% select ("std. ß[95% CI]", p))

options(knitr.kable.NA = "")
table_A3 <- kable (table_associations_long_2,  col.names = c("z-scores", "n obs", "n participants", "std. ß [95% CI]", "p", "std. ß [95% CI]", "p", "std. ß [95% CI]", "p", "std. ß [95% CI]", "p"), caption = "**Table A3: Interaction effects of plasma Aß42/40, p-Tau 181, GFAP and NfL over time on changes of parameters of cognition, structural and functional MRI and MRS.** Positive effects can be interpreted as stronger changes of the AD-related outcome over time, while negative effects explain a weakening of the effects over time. The interaction effects do not contain any information on the directionality of the effects of plasma biomarker concentration on MR-based parameters and have to be interpreted together with the main effect, which is why we provide effects for discrete changes of plasma biomarker concentration in the main text + table A4.", align = "c") %>%
  kable_classic() %>%
  add_header_above(c("  " = 3, "**Aß 42/40 x year**" = 2, "**p-Tau 181 x year**" = 2, "**GFAP x year**" = 2, "**NfL x year**" = 2))%>%
  footnote("The effects [95% CI] were estimated by linear mixed models adjusted for age, sex and education. Models including the MRS markers Myo-inositol and NAA were additionally weighted based on a measure of uncertainty (CRLB). All plasma biomarkers, MR-based and cognition parameters were z-standardized in order to allow for comparisons. *Abbreviations: Aß = amyloid beta, CI = confidence interval, CRLB = Cramer-Rao Lower Bound, DMN = Default Mode Network, GFAP = glial fibrillary acidic protein, NAA = N-Acetylaspartic acid, NfL= neurofilament light chain, NMM = NeuroMET Memory Metric, obs. = observations, part. = participants, p-Tau 181 = tau phosphorylated at threonine 181, Sal = Salience network. *",
           general_title = "")
  # %>%
  # save_kable("results/table_A3.html")
table_A3
```
